From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-29 Pi+/paper/reversible1.tex
denotational semantics in sec. 3.2

%%%%%%%%%
\subsection{Denotational Semantics}

\noindent The accompanying code includes a simple operational semantics for $\Pi$. A denotational semantics can be
specified by mapping each type to a finite set and each combinator to a bijection between finite sets, that is, a
permutation. In this section, we outline a simple semantics expressed in the metalanguage of set theory. This semantics
is formalized and extended to the metalanguage of homotopy type theory in Sec.~\ref{sec:reversibletwo}.

As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
we have that $\sizet{A} = \sizet{B}$. Hence the denotation of a type $A$ with $n$-elements will be a finite set
$\Fin[n]$ and the denotation of a combinator $c : A \iso B$ such that $\sizet{A} = \sizet{B} = n$ will be a function
from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. In more detail, we enumerate the values $v : A$ where
$\sizet{A} = n$ in a canonical order such that each value $v$ is given an index in $\{ 0,\cdots,n-1 \}$. The enumeration
is defined as follows:
\[\begin{array}{rcl}
    \mathit{enum}(\zerot) &=& [~] \\
    \mathit{enum}(\onet) &=& [ ~\Acon{tt}~ ] \\
    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
\end{array}\]
\noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
operation that appends all the subsequences in a sequence. For example, the 3 values $\inlv{\Acon{tt}}$,
$\inrv{(\inlv{\Acon{tt}})}$, and $\inrv{(\inrv{\Acon{tt}})}$ of type $\onet + (\onet + \onet)$ are given indices 0, 1, and 2
respectively. Given this canonical order, the combinator $\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$
would correspond to the permutation $(1~2~0)$ while the combinator $\assoclp : \onet + (\onet + \onet) \iso (\onet +
\onet) + \onet$ at the same type would correspond to the identity permutation.

>>>>

%%%%%%%%%
\subsection{Denotational Semantics (Outline)}

\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be
specified by mapping each type to a finite set and each combinator to a bijection between finite sets, that is, a
permutation. In this section, we outline a simple semantics expressed in the metalanguage of sets and functions. This
semantics will be formalized and extended to the metalanguage of homotopy type theory in Sec.~\ref{sec:reversibletwo}.

As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
$\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
\[\begin{array}{rcl}
    \denot{\onet + (\onet + \onet)} &=& \{ 0,1,2 \} \\
    \denot{(\onet + \onet) + \onet} &=& \{ 0,1,2 \} \\
    \\
    \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (0~1~2) \\
    \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (2~0~1)
\end{array}\]
where we have used the one-line notation for permutations with $(a~b~c)$ representing the
permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
enumeration of the elements of each type:
\[\begin{array}{rcl}
    \mathit{enum}(\zerot) &=& [~] \\
    \mathit{enum}(\onet) &=& [ ~\Acon{tt}~ ] \\
    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
\end{array}\]
\noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
operation that appends all the subsequences in a sequence.

Using the definition, we have:
\[\begin{array}{rcl}
    \mathit{enum}(\onet + (\onet + \onet)) &=& [ \inlv{\Acon{tt}},\inrv{(\inlv{\Acon{tt}})}, \inrv{(\inrv{\Acon{tt}})} ] \\
    \mathit{enum}((\onet + \onet) + \onet) &=& [ \inlv{(\inlv{\Acon{tt}})},  \inlv{(\inrv{\Acon{tt}})}, \inrv{\Acon{tt}} ]
  \end{array}\]
Thus, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of $\inlv{\Acon{tt}}$ which is 0 and
produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value $\inlv{(\inlv{\Acon{tt}})}$. Similarly,
$\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of $\inlv{\Acon{tt}}$ which is 0 and produces
index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value $\inrv{\Acon{tt}}$.
